---
title: "R Control Flow Statements"
subtitle: "Building Blocks for Automated Decision-Making"
author: "Sherine Paul Arthur | 401147341 | paul_arthur.sherine@stud.hs-fresenius.de"
date: today
format:
  revealjs:
    theme: simple
    slide-number: true
    code-line-numbers: true
    highlight-style: github
    self-contained: true
    smaller: true
    scrollable: false
    footer: "Data Analysis for Decision-Making (SS 2026) | Prof. Dr. Stephan Huber"
  pdf:
    documentclass: beamer
    pdf-engine: pdflatex
---

## Where we left off

we got the data into R. and we saw how to describe it. so whats next?

a data frame sitting in your environment doesn't do anything on its own.

At some point you need your code to **make decisions** — check a condition, repeat a task, skip a row that doesn't belong.

That is what we are covering today.

---

## What is control flow?

> "Control flow refers to the order in which the instructions of a program are executed."
>
> — Lenovo Tech Glossary

By default R runs every line, top to bottom, no questions asked.

Control flow gives you three tools to change that:

- **Branching** — run code only when a condition is met
- **Looping** — repeat code automatically
- **Jumping** — skip or stop mid-loop



---

## Branching — before we write any code

Imagine you have the final exam scores for everyone in this room.

- Score ≥ 90 → excellent
- Score 60–89 → passed
- Score < 60 → needs follow-up

You would not check each one manually. You write the rule once and let R apply it to every row.

That rule is an `if / else` statement.

---

## `if` / `else`

```{r}
#| echo: true
score <- 85

if (score >= 90) {
  print("Excellent")
} else if (score >= 60) {
  print("Passed")
} else {
  print("Needs follow-up")
}
```

R checks each condition top to bottom and stops at the first `TRUE`.

>  Inside `if()` always use `==` to compare, never `=`
> `=` stores a value — `==` asks a question

---

## `ifelse()` — the same logic, for an entire column

`if / else` works on **one value**. Your data has hundreds of rows.

```{r}
#| echo: true
# Final exam results
scores  <- c(Ali=82, Luisa=47, Roopam=91, Mahdi=55, Cagan=68)

result  <- ifelse(scores >= 60, "Pass", "Fail")
print(result)
```

`ifelse()` applies the condition to every element at once.

Same logic as `if / else` — completely different scale.

---

## Looping — the idea

we know how to get the mean of one column.

but what if you have 20 columns and need the mean of each?

```{r}
#| echo: true
data <- data.frame(
  sales    = c(200, 150, 300),
  expenses = c(100, 120, 90),
  profit   = c(100, 30, 210)
)

for (col in names(data)) {
  cat("Mean of", col, ":", mean(data[[col]]), "\n")
}
```

One loop. Works for 3 columns or 300.

---

## `for` loops

Go through every item in a sequence and do something with each one.

```{r}
#| echo: true
# Weekly coffee budget per student (€)
coffee <- c(Sherine=18, Milad=22, Nwanneka=9, Mohammed=31)

for (name in names(coffee)) {
  if (coffee[name] > 25) {
    cat(name, "— needs an intervention\n")
  } else {
    cat(name, "— doing fine\n")
  }
}
```

The loop variable takes each value one at a time. When it runs out — it stops.

---

## `while` loops

Use when you don't know how many steps it will take.

```{r}
#| echo: true
balance <- 500
month   <- 0

while (balance < 1000) {
  balance <- balance * 1.05   # 5% monthly return
  month   <- month + 1
}

cat("Reached €1000 after", month, "months\n")
```

Keeps going as long as the condition is `TRUE`.
Stops the moment it becomes `FALSE`.

---

## The infinite loop

If the condition never becomes `FALSE` — R freezes.

```r
# DO NOT RUN THIS
x <- 1
while (x > 0) {
  print(x)
  # x never changes — this never stops
}
```

Always make sure something inside the loop **moves you toward the exit**.

In the savings example — `balance` keeps growing until the condition fails.

---


## Jumping — interrupting a loop

Sometimes a loop needs to stop early or skip a step.

R gives you two tools for this: `break` and `next`.

---


## `break` and `next`

**`break`** — exit the loop immediately:

```{r}
#| echo: true
submissions <- c(Saketh=4.2, Adineh=3.8, Shravanthi=-99, Bahadori=5.1)

for (name in names(submissions)) {
  if (submissions[name] < 0) {
    cat("Missing submission from", name, "— stopping.\n")
    break
  }
  cat(name, "— submitted OK\n")
}
```

**`next`** — skip this one, keep going:

```{r}
#| echo: true
for (name in names(submissions)) {
  if (submissions[name] < 0) next
  cat(name, "— submitted OK\n")
}
```

---

## Putting it all together

```{r}
#| echo: true
# -999 is a common placeholder for missing data
raw   <- c(320, -999, 415, 280, -999, 390, 610, 305)
clean <- c()

for (x in raw) {
  if (x == -999) next    # skip missing
  if (x > 500)   next    # skip outliers
  clean <- c(clean, x)
}

cat("Original:", length(raw), "values\n")
cat("Clean:   ", length(clean), "values\n")
cat("Mean:    ", round(mean(clean), 1), "\n")
```

A `for` loop + `if` + `next` — three tools, one clean dataset.

---

## Summary

- `if / else` — one decision on one value
- `else if` — more than two possible outcomes
- `ifelse()` — decision across a whole vector
- `for` — repeat over a known sequence
- `while` — repeat until a condition changes
- `break` — exit the loop immediately
- `next` — skip this iteration, keep going

---

## Your turn — exercise

You have this vector:

```r
numbers <- c(3, -1, 7, 0, -4, 9, 2, -8, 5, 1)
```

Write a `for` loop that prints `"positive"`, `"negative"`, or `"zero"` for each number.

**Bonus:** use `next` to skip any number greater than 6.


---

## Solution

```{r}
#| echo: true
numbers <- c(3, -1, 7, 0, -4, 9, 2, -8, 5, 1)

for (n in numbers) {
  if (n > 6) next

  if (n > 0) {
    cat(n, "→ positive\n")
  } else if (n < 0) {
    cat(n, "→ negative\n")
  } else {
    cat(n, "→ zero\n")
  }
}
```

---

## References

Grolemund, G. (2014). *Hands-on programming with R*. O'Reilly. https://rstudio-education.github.io/hopr/

Hamelg. (2021). *Intro to R part 11: Control flow*. Kaggle. https://www.kaggle.com/code/hamelg/intro-to-r-part-11-control-flow

Lenovo. (n.d.). *Control flow*. https://www.lenovo.com/us/en/glossary/control-flow/

Peng, R. D. (2020). *R programming for data science*. Leanpub. https://bookdown.org/rdpeng/rprogdatascience/

Phillips, N. D. (2018). *YaRrr! The Pirate's Guide to R*. https://bookdown.org/ndphillips/YaRrr/

R Core Team. (2024). *R: A language and environment for statistical computing*. https://www.R-project.org/

Wickham, H. (2019). *Advanced R* (2nd ed.). Chapman and Hall. https://adv-r.hadley.nz/

---

## Access the materials

Scan to open the presentation and handout on your device:

**Presentation:**
https://sherinepaularthur.github.io/presents/control_flow_presentation.html 

**Handout:**
https://sherinepaularthur.github.io/presents/control_flow_handout.html
