---
title: "R Control Flow Statements"
subtitle: "Building Blocks for Automated Decision-Making"
author: "Sherine | 401147341 | paul_arthur.sherine@stud.hs-fresenius.de"
date: today
format:
  revealjs:
    theme: simple
    slide-number: true
    code-line-numbers: true
    highlight-style: github
    self-contained: true
    smaller: true
    scrollable: false
  pdf:
    documentclass: beamer
---

## Where we left off

Ali got the data into R. Luisa showed you how to read it.

But data sitting in a data frame doesn't do anything on its own.

At some point you need to make decisions with it — check conditions, repeat tasks, skip things that don't belong.

That's the step we're covering today.

---

## What is control flow?

> "Control flow refers to the order in which the instructions of a program are executed."
> — Lenovo Tech Glossary

By default, R runs every line of your code, top to bottom, no questions asked.

Control flow gives you three tools to change that:

| Tool | What it does |
|---|---|
| **Branching** | Run code only when a condition is met |
| **Looping** | Repeat code automatically |
| **Jumping** | Skip or stop mid-loop |

---

## Branching — the idea

Imagine checking grades for 300 students:

- Score ≥ 90 → send congratulations  
- Score 60–89 → send pass confirmation  
- Score < 60 → flag for follow-up  

You wouldn't do this manually. You write the rule once — R applies it to every row.

That rule is an `if / else` statement.

---

## `if` / `else`

```{r}
#| echo: true
score <- 85

if (score >= 90) {
  print("Excellent — send congratulations")
} else if (score >= 60) {
  print("Passed — send confirmation")
} else {
  print("Failed — flag for follow-up")
}
```

R checks each condition top to bottom and stops at the first `TRUE`.

> ⚠️ Always use `==` to compare inside `if()`, never `=`  
> `=` stores a value — `==` asks a question

---

## `else if` — more than two outcomes

Chain as many conditions as you need. R works through them in order.

```{r}
#| echo: true
temperature <- 22

if (temperature >= 35) {
  print("Extreme heat")
} else if (temperature >= 25) {
  print("Warm")
} else if (temperature >= 15) {
  print("Mild")
} else {
  print("Cold")
}
```

Think of it as a checklist — R stops the moment it finds a match.

---

## `ifelse()` — decisions across a whole vector

`if / else` works on one value. Your data has thousands of rows.

`ifelse()` applies the condition to every element at once.

```{r}
#| echo: true
scores <- c(45, 72, 88, 55, 91, 60)

result <- ifelse(scores >= 60, "Pass", "Fail")
print(result)
```

Same logic as `if / else` — completely different scale.

---

## Looping — the idea

Luisa showed you the mean of one column.

What if you have 20 columns?

```{r}
#| echo: true
data <- data.frame(
  sales    = c(200, 150, 300),
  expenses = c(100, 120, 90),
  profit   = c(100, 30, 210)
)

for (col in names(data)) {
  cat("Mean of", col, ":", mean(data[[col]]), "\n")
}
```

One loop. Works for 3 columns or 300.

---

## `for` loops

Go through every item in a sequence and do something with each one.

```{r}
#| echo: true
products <- c("Laptop", "Phone", "Tablet")

for (item in products) {
  cat("Processing:", item, "\n")
}
```

`item` takes each value from the vector one at a time.  
When the vector runs out — the loop stops automatically.

---

## `while` loops

Use when you don't know how many steps it will take.

```{r}
#| echo: true
budget <- 1000
spent  <- 0
items  <- 0

while (spent + 150 <= budget) {
  spent <- spent + 150
  items <- items + 1
}

cat("Bought", items, "items | Spent €", spent, "\n")
```

Keeps going as long as the condition is `TRUE`.  
Stops the moment it becomes `FALSE`.

---

## ⚠️ Infinite loops

If the condition never becomes `FALSE` — R freezes.

```r
# DO NOT RUN
x <- 1
while (x > 0) {
  print(x)
  # x never changes — runs forever
}
```

Always make sure something inside the loop moves you toward the exit.

In the budget example — `spent` keeps increasing until the condition fails.

---

## `break` and `next`

**`break`** — exit the loop immediately:

```{r}
#| echo: true
values <- c(12, 7, 33, -5, 9)

for (v in values) {
  if (v < 0) { cat("Negative found:", v, "\n"); break }
  cat("OK:", v, "\n")
}
```

**`next`** — skip this one, keep going:

```{r}
#| echo: true
for (v in values) {
  if (v < 0) next
  cat(v, "")
}
```

---

## Putting it all together

```{r}
#| echo: true
data  <- c(200, -999, 450, -999, 310, 180)
clean <- c()

for (x in data) {
  if (x == -999) next   # skip missing
  if (x > 400)   next   # skip outliers
  clean <- c(clean, x)
}

cat("Cleaned:", clean, "\n")
cat("Mean:   ", mean(clean), "\n")
```

A `for` loop + `if` + `next` — three tools cleaning a dataset automatically.

---

## Summary

| Statement | When to use it |
|---|---|
| `if / else` | One decision on one value |
| `else if` | More than two possible outcomes |
| `ifelse()` | Decision across a whole vector |
| `for` | Repeat over a known sequence |
| `while` | Repeat until a condition changes |
| `break` | Exit the loop immediately |
| `next` | Skip this iteration, keep going |

---

## Exercise

You have this vector:

```r
numbers <- c(3, -1, 7, 0, -4, 9, 2, -8, 5, 1)
```

Write a `for` loop that prints `"positive"`, `"negative"`, or `"zero"` for each number.

**Bonus:** use `next` to skip any number greater than 6.

*Open a new R script — you have 7 minutes.*

---

## Solution

```{r}
#| echo: true
numbers <- c(3, -1, 7, 0, -4, 9, 2, -8, 5, 1)

for (n in numbers) {
  if (n > 6) next

  if (n > 0) {
    cat(n, "→ positive\n")
  } else if (n < 0) {
    cat(n, "→ negative\n")
  } else {
    cat(n, "→ zero\n")
  }
}
```

---

## References

Grolemund, G. (2014). *Hands-on programming with R*. O'Reilly. https://rstudio-education.github.io/hopr/

Hamelg. (2021). *Intro to R part 11: Control flow*. Kaggle. https://www.kaggle.com/code/hamelg/intro-to-r-part-11-control-flow

Lenovo. (n.d.). *Control flow*. https://www.lenovo.com/us/en/glossary/control-flow/

Peng, R. D. (2020). *R programming for data science*. Leanpub. https://bookdown.org/rdpeng/rprogdatascience/

R Core Team. (2024). *R: A language and environment for statistical computing*. R Foundation. https://www.R-project.org/

Wickham, H. (2019). *Advanced R* (2nd ed.). Chapman and Hall. https://adv-r.hadley.nz/

---

## Affidavit

I hereby affirm that this submitted paper was authored unaided and solely by me. No other sources than those in the reference list were used. Parts taken verbatim or analogously from other works have been properly cited. This paper has not been submitted to any other examination board and has not been published.

I acknowledge that the university may use plagiarism detection software to check my submission. I agree to cooperate with any investigation of suspected plagiarism.

---

## Checklist

- [x] Quarto files submitted
- [x] HTML files submitted
- [x] PDF files submitted
- [x] Title page contains name, email, matriculation number
- [x] Bibliography in APA style
- [x] R code included
- [x] Exercise included
- [x] Affidavit filled out
- [ ] Published on GitHub
- [ ] GitHub link shared with audience

[Sherine] | [Date] | [Cologne]
